Ribosome accumulation during early phase resistance training in humans

### 2022-01-17 response-to-reviews-rev1.Rmd

---
title: "APH-2021-12-0652: Response to the reviewers"
author: "Daniel Hammarström"
date: "2022-01-17"
output: pdf_document
---

## Referee 1

> This manuscript investigates the induction of ribosome biogenesis and accumulation of RNA/ribosomes during the early phase of resistance training. Subjects performed unilateral RE bilaterally but with differing loading protocols. One leg trained with a constant volume whereas the other trained with variable volume. Control subjects were biopsied at relevant time points. The manuscript presents evidence in favor of that skeletal muscle in the naïve state is very responsive with respect to ribosome biogenesis following acute RE. With repeated training sessions this induction is maintained but not further stimulated. Most of the effect on total RNA was seen during the first training block. UBF levels were related to RNA accumulation and rate of total RNA accumulation predicted RT-induced muscle hypertrophy. With eight days of detraining, around 20% of total RNA was lost, indicating that the elevated RNA pool needs to be maintained not to decrease.
The manuscript is highly interesting and ambitious in its design. I think that it adds to the published literature and could potentially be a future important reference piece. However, I have some concerns and comments that I’d like for the authors to address, as I think that they can improve the manuscript further.

> The western blot membranes in the paper does not really translate into the presented data. In my opinion the bands look rather faint and equal in intensity across the time course. Could the authors perhaps provide raw data and membranes for RPS6 and UBF just for the reviewers?

> Is pre-rRNA 47S ETS and 45S ETS per unit tissue weight a previously reported way to present qPCR data? Does the data look the same in relation to 1 or more house keeping genes?

One transcribed pre-rRNA results in a single end product, one ribosome and in the context of protein synthesis the amount of ribosomes per cell or tissue mass is highly correlated with protein synthesis (Milward 1976, West 2016). In contrast to rRNA processing, mRNA transcription and translation could to some degree be seen as competitive between different mRNA.

This translates to different normalization strategies. The use of a houskeeping gene when investigating ribosomal biogenesis is not interesting as it does not

> Could the authors please comment on the apparent disconnect between increase in RNA and RPS6 protein levels?

> The concept that the increase in RNA content predicts subsequent hypertrophy is very interesting. Even more intriguing is the finding that total RNA mid training negatively correlates to muscle hypertrophy is fascinating! I would encourage the authors to deepen the discussion on these key data points, especially in the light of more and more evidence in favor of ribosome specialization. Could a high percentage of new ribosomes be advantageous for growth in more ways than just contributing greater numbers?

The referee raises an interesting point

> Please comment on that UBF levels where maintained at the detraining time point but total RNA decreasing. Would it be relevant to analyze UBF phosphorylation?

> The fact that RNA decreased at the detraining time point is interesting and indicates the importance of actively maintaing translational capacity not to loose it. I think the discussion could benefit from including a short section of the role for ribophagy in this setting?

> The methods in the abstract must be improved. In its current form it lacks key information such as training duration, the number of controls are states as n=7 but n=8 in the M&M section and I’d like for the authors to try and specify biopsy time points.

> Page 3, line 48-49. I encourage the authors to consider including an additional reference (Figueiredo et al 2021 J Phys). This paper shows the specific response of ribosome biogenesis in response to RE and not Endurance type exercise.

> Page 4, line 8: Please clarify “numberical lowering”. Could this be rephrased?

> Page 4, line 10: Double “..”

> Page 4, line 15: Double “..”

> Page 4, line 26: Should “to” be removed?

> Page 11, line 20: n=7 or 8? Please check abstract

> Page 11, line 35: Should “a” be removed? Please double check if it’s 7 or 8 days of detraining. Abstract states 8.

> Page 12, line 19: please clarify “leg being training in the rest period between sets in the first leg. Does this means that they training continuously without break?

> Page 13, line 52: 2-4 passes were made with the Bard Magnum 12-14 gauge needles but only 1-2 aliquots of the samples were frozen. What happened to the other material? If I misunderstood, can the text be rewritten for clarity?

> Page 14, line 8: Was the spike-in used to normalize RNA extraction efficiency? If yes, please state this clearly.

> Page 14, line 49: Does “normalized” refer to denatured?

> Page 15, line 15: Does the authors refer to amplicon size when they write primer sizes?

> Page 31, Figure legends: I think that the figure legends could benefit from a bit more substance. For example Fig 1A, perhaps indicate bar color as black (CONST) and yellow (VAR)?

## Referee: 2

> The authors sought to perform a time course investigation of ribosome density and biogenesis markers after knee extensor training.  Phenotype variables were also assessed.  Training led to increases in muscle growth (VL thickness) as well as in total RNA, rRNA, UBF and rpS6.  Training volume did not play a role in these adaptations.  Interestingly, the rate of total RNA increases predicted hypertrophy.  Likewise, training cessation led to decrements in total RNA abundance.
I love this study.  This continues to contribute to our knowledge regarding the role of ribosome biogenesis in muscle hypertrophy, and is delicate human work with multiple time point biopsies.
I have minor comments in order of appearance, and congratulate the authors on their fine work.

>  ABSTRACT: Reads well and concise.  Nice work, and no comments.

> INTRODUCTION 
- Line 7, minor: “promotes” should read as “promote”
- Line 14, minor: “with subsequent repeated bouts” should read as “and subsequent repeated bouts”
- Line 18, minor: “resting synthetic rate of muscle protein” better reads as “basal muscle protein synthesis rates”
- Lines 48-50, minor: “bouts lead to accumulation of mature rRNA thus also total RNA and presumably functional ribosomes”…this sentence is long, and the word “thus” seems to make it a run-on.  I’d rephrase the sentence to read more coherently.
Page 4
- Lines 10 and 15, minor: each sentence has two stops.
- Line 33, minor: no need to use the phrase “per se”
RESULTS
- Line 51 (and throughout the results), clarification: is “9.2%-point difference” the same as “9.2% difference”?  If so, adjust and remove “point” from all descriptors as this is confusing.
- Other than this, the results, while dense and full of data, are well-written.
DISCUSSION
Page 7
- Line 17, minor: again, “per se” can be removed.
- Line 19, minor: “was not affected” better reads as “were not affected”
- Line 39, minor: “session” should read as “sessions”

> I urge the authors to be creative somewhere in the discussion by perhaps adding a paragraph after the one found in lines 29-56.  We’ve done a lot of work in the ribosome biogenesis area, and we’ve thought a good bit about total RNA/unit muscle with regard to what it represents.  The authors correctly articulated this as a proxy of ribosome density.  However, as muscle tissue and myofibers volumetrically increase, this will affect the metric (the same holds true with mitochondrial biogenesis and using CS activity as a proxy of mitochondrial volume density).  An easy way to interpret this is in the following example:
o Muscle tissue increases 5% in thickness with RT (and myofibers scale accordingly as well)
o Total RNA/unit muscle does not change in value
Most would interpret this as ribosome biogenesis not occurring.  However, biogenesis likely occurred because, as muscle grew, ribosome biogenesis scaled with growth thus leading to no change in RNA concentration.  Our laboratory discusses this important concept with regard to RT and mitochondrial biogenesis (PMID: 32162291, PMID: 34646153).

> Anyhow, this is no critique of the authors’ hard and excellent work.  However, I think the authors re-iterating this point as a stand-alone paragraph would benefit readership.  In addition, I believe taking a moment to define what the total RNA/ribosome density metric conceptually represents, and then (in this newly added paragraph) briefly walking the reader through each phase of the data, and what could be going on with biogenesis in the context of a growing muscle would be interesting to see.  For instance, in Figure 3G, my interpretation (much like the authors’) is that ribosome biogenesis is very rapid at the onset of RT where little functional growth is likely occurring.  And like the authors, I believe this likely was needed for eventual growth to occur.  However, the “no change” in RNA concentrations session 9-12 as muscle is continuing to grow may not indicate a plateau in biogenesis given that this is the time during which muscle growth is likely starting to ramp up.  Hence, biogenesis rates could be equally as ramped here compared to sessions 1-9 of RT.

METHODS

Page 11

- Lines 9-19: Authors stated that 18 participants were recruited, but then go on to say that n=11 TRAIN and n=8 CTRL completed.  Please reconcile.
- Line 35: Authors stated 7 days after detraining, but in the discussion (page 7, line 19) they mention after 8 days of detraining.  Please reconcile.
Page 12
- Line 39, minor: “restitution” should read as “rest”.  Please change accordingly, and if used throughout the paper, please alter.
Page 15
- Line 36, minor: “Glycin” should read as “Glycine”.  In addition, and molar unit is needed.

### 2022-01-18 response-to-reviews-rev1.Rmd

## Referee 1

> This manuscript investigates the induction of ribosome biogenesis and accumulation of RNA/ribosomes during the early phase of resistance training. Subjects performed unilateral RE bilaterally but with differing loading protocols. One leg trained with a constant volume whereas the other trained with variable volume. Control subjects were biopsied at relevant time points. The manuscript presents evidence in favor of that skeletal muscle in the naïve state is very responsive with respect to ribosome biogenesis following acute RE. With repeated training sessions this induction is maintained but not further stimulated. Most of the effect on total RNA was seen during the first training block. UBF levels were related to RNA accumulation and rate of total RNA accumulation predicted RT-induced muscle hypertrophy. With eight days of detraining, around 20% of total RNA was lost, indicating that the elevated RNA pool needs to be maintained not to decrease.
The manuscript is highly interesting and ambitious in its design. I think that it adds to the published literature and could potentially be a future important reference piece. However, I have some concerns and comments that I’d like for the authors to address, as I think that they can improve the manuscript further.

> The western blot membranes in the paper does not really translate into the presented data. In my opinion the bands look rather faint and equal in intensity across the time course. Could the authors perhaps provide raw data and membranes for RPS6 and UBF just for the reviewers?

```{r}
## Raw data of western blot membranes

western_data <- readRDS("./data/derivedData/western-compile/western_data.RDS")
# Save data 
western_data <- readRDS("./data/derivedData/western-compile/western_data_calibrated.RDS")

western_data %>%
  mutate(detrain = if_else(time == "post1w", "detrain", "train"), 
         time = if_else(time %in% c("post1w", "postctrl"), "12", 
                        if_else(time == "S1c", "1", 
                                gsub("S", "", time))), 
         group = if_else(cond == "ctrl_leg", "ctrl", "train"), 
         time = as.numeric(time), 
         time.fact = as.factor(time)) %>%

  group_by(target, participant, group, time.fact, detrain) %>%
  summarise(expression = mean(expression)) %>%

  ggplot(aes(time.fact, expression, color = group)) + 
  geom_violin() + 
  geom_jitter(height = 0, width = 0.1) +
  facet_wrap(~ target)

```

### 2022-02-01 manuscript.Rmd
updates revision

---
title: "Ribosome accumulation during early phase resistance training in humans"
author: Daniel Hammarström^1,2,£^*, Sjur J. Øfsteng^2,£^*, Nicolai B. Jacobsen^1^, Krister B. Flobergseter^1^, Bent R. Rønnestad^1^, Stian Ellefsen^1,3^
date: ''
output:
  officedown::rdocx_document:
    reference_docx: resources/manuscript_word_template.docx
csl: resources/acta-physiologica.csl
bibliography: resources/bibliography.bib
header-includes: 
  \usepackage{wasysym}
editor_options:
  chunk_output_type: console
---
<!-- This is a comment  -->
<!-- Add comments after each reference for tracking (eg[@RN2364]<!-- comment -->
<!-- After references, add a row break without space (this will put the reference on the end on the line). -->

@@

**Methods:** A training group (n=11) performed unilateral knee extension with constant and variable volume (6 and 3-9 sets per session respectively) allocated to either leg. Vastus lateralis biopsies obtained at rest and 48 hrs after selected training sessions and after eight days of de-training and from a non-training control group (n=7) were assayed for ribosome abundance and biogenesis markers. In addition, muscle thickness was measured before and after training and control periods.

@@

The observed rates of RNA accumulation over the entirety of the intervention were found to be a determinant of changes in muscle thickness (after controlling for average total RNA levels).
Individuals with higher rates of accumulation showed larger accretion of muscle mass.
This supports the notion that ribosomal biogenesis is an important determinant of RT-induced muscle hypertrophy, with previous studies showing that increases in total RNA are positively correlated with increases in muscle mass 
[@RN1644;<!-- Figueriedo  -->
@RN1897;
@RN2055],<!-- Mobely Three groups, no clear group * time in RNA incrinteraction but 32 vs 8 % increase in high vs low responders, although correlation RNA incr with MT incr -->
<!-- Reidy correlation between total RNA and muscle hypertrophy-->
differs between individuals displaying low vs. high levels of muscle hypertrophy in response to RT
[@RN1755]<!-- Stec 2016  -->
an contribute to explain RT volume-dependent changes in muscle mass and strength
[@RN2358].
In addition, supression of ribosomal biogenesis in *in vitro* models leads to halted muscle cellular growth in some 
[@RN1632;<!-- Nader 2005 blockage of mTOR halts ribo bio and growth -->
@RN1754<!-- West 2016 Inhibition of c-Myc (through inibitor) reduces RNA and MPS-->;
@RN1755]<!-- Stec specific pol I inhibition -->
but not all studies
[@RN1929].<!-- Crossland 2017 inhibition of ribo bio did not limoit igf-1 stimulated growth  -->
Conversely, individual variation in fixed amounts of total RNA was not found to determine muscle mass accretion, and higher levels of total RNA was instead associated with a tendency towards lowered muscle growth.
Overall, the rate of increases in ribosomal density thus seems to be a better predictor of individual RT-induced changes in muscle mass than absolute ribosomal density, suggesting that net increases in ribosomal biogenesis may be a core determinant of RT responsiveness.
Interestingly, the interaction between rRNA synthesis rate and muscle mass accretion (but not between ribosomal content and muscle mass accretion) may shed light on observed differences in muscular responses to RT between young and old individuals.
Whereas aged muscle display higher levels of total RNA at rest 
[@RN1656]<!-- Stec 2015 higher levels of RNA/mg in old -->
they show reduced changes in total RNA levels in response to RT
[@RN1809],<!-- Brook 2016 increase in RNA per mg not seen in old, data at baseline collected at different timepoints? -->
potentially explaining their alleged poorer overall hypertrophic responses [@RN1809].
Whether these cellular characteristics are related to e.g. differences in fiber type distributions [@RN1707] remains to be determined.
Together, these results and perspectives emphasizes on the potentially crucial role of RT-induced ribosomal synthesis for adaptations to training, making ribosomal responses to RT an interesting biomarker in relation to manipulation of training loads for specific populations.

@@

After eight days of de-training, total RNA and rRNA levels per weight unit muscle tissue returned toward baseline levels, though without concomitant reversal of muscle thickness, which remained at elevate levels.
This was likely caused by attenuated rRNA transcription, a notion that was supported by reversal of pre-rRNA abundances and possibly by lowered UBF protein levels, though this was not confirmed as statistically robust.
This supports the idea that ribosomal biogenesis is a cellular activity on demand, possibly relating to its relative expense [@RN1920] also in muscle tissue.
Based on this notion, and the fact that RT volume is known to be a potent modulator of molecular mechanisms determining protein synthesis and ribosomal biogenesis including induction of c-Myc expression, mTOR activation
[@RN2358; <!-- Hammarström volume affects ribo biogenesis and mTOR and c-myc -->
@RN1837; <!-- Ahtianinen 2015 volume affects mTOR differently -->
@RN791]<!-- Burd ex volume protein synthesis and anabolic [rpS6] phospho signalling-->,
subsequent total RNA increases [@RN2358] and post exercise protein synthesis [@RN791]
and subsequent training outcomes [@RN1767; @RN2358], we hypothesized that fluctuations in training volume would be reflected in markers of ribosomal biogenesis.
When comparing VAR to CONST in the present study we found only one part of the pre-rRNA, 45S ETS, to be differentially expressed and only so after Session 12 in favor of VAR together with a tendency towards rescued UBF levels after de-training in response to increased volume in the VAR but not CONST protocol.
These observations do not give support to a clear effect of fluctuations in training volume on total RNA levels or rRNA expression within a relatively short and training-intensive intervention, though it should be noted that the time point with increased 45S ETS expression was preceded by a period of increased training volume, suggesting a potential interaction between time and volume. 
Indeed, both training protocols utilized in the present study increased muscle strength and induced muscle hypertrophy to a similar degree.
From a general perspective, albeit volume is an important determinant of increases in muscle strength and mass
[@RN1767;<!-- schoenfeld 2016 dose response  -->
@RN2492],<!-- Ralston 2017 Weekly sets volume on muscle strength  -->
differences in organization of training loads is likely of minor importance when training volumes are equated over time
[@RN2572].<!-- Grgic 2017 volume equated periodization schemes led to similar muscle hyeprtrophy -->
It is important to note that RT in the current study was performed with the same volume in the first four sessions, something that could have been more than enough to maximize rRNA transcription in previously untrained individuals.
This is supported by the observation that pre-rRNA increased rapidly initially in both protocols with minimal changes in response to subsequent sessions, regardless of exercise volume.
The CONST protocol in the present study corresponded to volumes used in the moderate volume condition in a previous study from our lab (three sets in two exercises activating knee extensor muscles) [@RN2358].
There, higher levels of total RNA were observed after four sessions in the moderate compared to a low volume protocol [@RN2358].
Interestingly, using a progressive volume protocol in well-trained participants, increases in total RNA have been reported throughout six weeks of training
[@RN2225].<!-- Haun total RNA increase througout a 6 week intervention -->
Altough this observation was done in well-trained participants performing a high volume protocol without a control condition with constant volume, compared to constant volume protocols
[@RN1809;<!-- Brook 2016 -->
@RN2358; and the present study],<!-- Hammarström 2020 -->
progressive volume may thus increase ribosomal abundance to a higher degree and provide a measure to avoid the plateau phase seen in the present study.

@@

## Muscle strength, body composition and muscle thickness assessments
Muscle strength was assessed as maximal voluntary isokinetic (90&deg; sec<sup>-1</sup>) and isometric (60&deg; angle, fully extended leg 0&deg;) knee extension torque. After a brief warm-up (5-min cycling, RPE 12-14), participants were seated and secured in the individually adjusted dynamometer. Participants were instructed to gradually increase their effort during three warm-up repetitions (50, 60 and, 70% of subjective maximal effort). After a 30-sec restitution period participants were instructed to perform three repetitions with maximal effort in the concentric phase. Sixty seconds after the isokinetic test the lever automatically moved to a 60&deg; angle and participants were instructed to push against the lever enough to see feedback from the visual feedback system. After an additional 15-sec restitution period, participants were instructed to push against the lever with maximal effort. Within the same assessment session, participants remained seated in the dynamometer for measurement perfomed on both legs. The first measurement was alternated between legs every other session. For statistical treatment of the data, all successful attempts were used. The last strength assessment at baseline was performed at least seven days prior to the first biopsy sampling. At least one of the baseline strength tests was performed on separate day with two sessions allowed to be perform on the same day with a short rest between assessments. Post training assessments were performed 48 hours and eight days after the last session.

@@

## Immunoblotting
Protein samples (20 &#956;g) were separated on 4-20% Tris-Glycin gels (Criterion TGX Precast Gels, Bio-Rad) at 250 V for 45 min using the recommended running buffer (25 mM Tris, 192 mM Glycin, 0.1% SDS). All samples from the same participant were run on the same gel and all samples were run in at least duplicates. When samples were compared between participants, signals were expressed per a calibrator sample from each participant that in turn was measured on a separate gel with all calibrator samples. Separated samples were transferred to &#956m PVDF membranes (Immun-Blot, Bio-Rad) using wet transfer (25 mM Tris, 192 Glycin, 10% vol/vol methanol) at a constant voltage of 300 mA for 3 h. Membranes were then stained to confirm transfer and enable total protein quantification using a reversible protein stain (Pierce Reversible Protein Stain, Thermo Fisher Scientiﬁc). Primary antibodies were acquired to detect UBF (F-9, sc-13125, Santa-Cruz Biotechnology, Dallas, Texas, USA) and ribosomal protein S6 (54D2, #2317, Cell Signaling Technology, Danvers, MA, USA). After blocking (Tris-buffered saline blocking buffer, 20 mM Tris, 150 mM NaCl, 5% fat-free milk, 0.1% Tween-20), membranes were incubated over-night with primary antibodies diluted in blocking buffer (UBF, 1:200; S6, 1:1000) followed by incubation with a secondary antibody conjugated to horseradish peroxidase (Anti-mouse IgG, #7076, Cell Signaling Technology, 1:10000). Membranes were washed 6 &#215; 5 min after incubation with primary antibodies and 8 &#215; 5 min after incubation with the secondary antibody. All incubation and washing steps were performed at 4&deg;C using an automatic membrane processor (BlotCycler, Precision Biosystems, Mansﬁeld, MA, USA). Chemiluminescent signals from membranes were detected after 5 min incubation in substrate (Super Signal West Femto Maximum Sensitivity Substrate, Thermo Fisher Scientiﬁc) using a documentation system. Total protein content was quantified from whole membrane images and defined as the mean gray value of the whole lane. Between-lane gray values were used as background subtracted from protein values. Total protein quantification was done using ImageJ Fiji [@RN2561]. Chemiluminescence signals were quantified using Image Studio Lite (LI-COR Biotechnology, Lincoln, NE, USA).

>>>>

---
title: "Ribosome accumulation during early phase resistance training in humans"
author: Daniel Hammarström^1,2,£^*, Sjur J. Øfsteng^2,£^*, Nicolai B. Jacobsen^1^, Krister B. Flobergseter^1^, Bent R. Rønnestad^1^, Stian Ellefsen^1,3^
date: ''
output:
  officedown::rdocx_document:
    reference_docx: resources/manuscript_word_template.docx
csl: resources/acta-physiologica.csl
bibliography:
  - resources/bibliography.bib
  - references.bib
header-includes: 
  \usepackage{wasysym}
editor_options:
  chunk_output_type: console
---
<!-- This is a comment  -->
<!-- Add comments after each reference for tracking (eg[@RN2364]<!-- comment -->
<!-- After references, add a row break without space (this will put the reference on the end on the line). -->

@@

**Methods:** A training group (n = 11) performed <span custom-style="CHANGE">12 sessions (3-4 sessions per week) of</span> unilateral knee extension with constant and variable volume (6 and 3-9 sets per session respectively) allocated to either leg. <span custom-style="CHANGE">Ribosome abundance and biogenesis markers were assessed from</span> vastus lateralis biopsies obtained at<span custom-style="CHANGE"> baseline and 48 hrs after sessions 1, 4, 5, 8, 9, 12</span> and after eight days of de-training and from a non-training control group (n = 8). In addition, muscle thickness was measured before and after training and control periods.

@@

The observed rates of RNA accumulation over the entirety of the intervention were found to be a determinant of changes in muscle thickness (after controlling for average total RNA levels).
Individuals with higher rates of accumulation showed larger accretion of muscle mass.
This supports the notion that ribosomal biogenesis is an important determinant of RT-induced muscle hypertrophy, with previous studies showing that increases in total RNA are positively correlated with increases in muscle mass 
[@RN1644;<!-- Figueriedo  -->
@RN1897;
@RN2055],<!-- Mobely Three groups, no clear group * time in RNA incrinteraction but 32 vs 8 % increase in high vs low responders, although correlation RNA incr with MT incr -->
<!-- Reidy correlation between total RNA and muscle hypertrophy-->
differs between individuals displaying low vs. high levels of muscle hypertrophy in response to RT
[@RN1755]<!-- Stec 2016  -->
an contribute to explain RT volume-dependent changes in muscle mass and strength
[@RN2358].
In addition, supression of ribosomal biogenesis in *in vitro* models leads to halted muscle cellular growth in some 
[@RN1632;<!-- Nader 2005 blockage of mTOR halts ribo bio and growth -->
@RN1754<!-- West 2016 Inhibition of c-Myc (through inibitor) reduces RNA and MPS-->;
@RN1755]<!-- Stec specific pol I inhibition -->
but not all studies
[@RN1929].<!-- Crossland 2017 inhibition of ribo bio did not limoit igf-1 stimulated growth  -->
Conversely, individual variation in fixed amounts of total RNA was not found to determine muscle mass accretion, and higher levels of total RNA was instead associated with a tendency towards lowered muscle growth.
Overall, the rate of increases in ribosomal density thus seems to be a better predictor of individual RT-induced changes in muscle mass than absolute ribosomal density, suggesting that net increases in ribosomal biogenesis may be a core determinant of RT responsiveness.
Interestingly, the interaction between rRNA synthesis rate and muscle mass accretion (but not between ribosomal content and muscle mass accretion) may shed light on observed differences in muscular responses to RT between young and old individuals.
Whereas aged muscle display higher levels of total RNA at rest 
[@RN1656]<!-- Stec 2015 higher levels of RNA/mg in old -->
they show reduced changes in total RNA levels in response to RT
[@RN1809],<!-- Brook 2016 increase in RNA per mg not seen in old, data at baseline collected at different timepoints? -->
potentially explaining their alleged poorer overall hypertrophic responses [@RN1809].
Whether these cellular characteristics are related to e.g. differences in fiber type distributions [@RN1707] remains to be determined.
<span custom-style="CHANGE">Furthermore, in light of accumulating evidence of tissue specific and heterogeneous ribosomal populations [@chaillou2019; @chaillou2015; @gupta2014], our results suggests a potential role of specialized newly synthesized ribosomes in muscle hypertrophy. Our results are however limited in terms of probing any such heterogeneity. </span>
Together, these results and perspectives emphasizes on the potentially crucial role of RT-induced ribosomal synthesis for adaptations to training, making ribosomal responses to RT an interesting biomarker in relation to manipulation of training loads for specific populations.

@@

After eight days of de-training, total RNA and rRNA levels per weight unit muscle tissue returned toward baseline levels, though without concomitant reversal of muscle thickness, which remained at elevate levels.
This was likely caused by attenuated rRNA transcription, a notion that was supported by reversal of pre-rRNA abundances and possibly by lowered UBF protein levels, though this was not confirmed as statistically robust.
This supports the idea that ribosomal biogenesis is a cellular activity on demand, possibly relating to its relative expense [@RN1920] also in muscle tissue.
Based on this notion, and the fact that RT volume is known to be a potent modulator of molecular mechanisms determining protein synthesis and ribosomal biogenesis including induction of c-Myc expression, mTOR activation
[@RN2358; <!-- Hammarström volume affects ribo biogenesis and mTOR and c-myc -->
@RN1837; <!-- Ahtianinen 2015 volume affects mTOR differently -->
@RN791]<!-- Burd ex volume protein synthesis and anabolic [rpS6] phospho signalling-->,
subsequent total RNA increases [@RN2358] and post exercise protein synthesis [@RN791]
and subsequent training outcomes [@RN1767; @RN2358], we hypothesized that fluctuations in training volume would be reflected in markers of ribosomal biogenesis.
When comparing VAR to CONST in the present study we found only one part of the pre-rRNA, 45S ETS, to be differentially expressed and only so after Session 12 in favor of VAR together with a tendency towards rescued UBF levels after de-training in response to increased volume in the VAR but not CONST protocol.
These observations do not give support to a clear effect of fluctuations in training volume on total RNA levels or rRNA expression within a relatively short and training-intensive intervention, though it should be noted that the time point with increased 45S ETS expression was preceded by a period of increased training volume, suggesting a potential interaction between time and volume. 
Indeed, both training protocols utilized in the present study increased muscle strength and induced muscle hypertrophy to a similar degree.
From a general perspective, albeit volume is an important determinant of increases in muscle strength and mass
[@RN1767;<!-- schoenfeld 2016 dose response  -->
@RN2492],<!-- Ralston 2017 Weekly sets volume on muscle strength  -->
differences in organization of training loads is likely of minor importance when training volumes are equated over time
[@RN2572].<!-- Grgic 2017 volume equated periodization schemes led to similar muscle hyeprtrophy -->
It is important to note that RT in the current study was performed with the same volume in the first four sessions, something that could have been more than enough to maximize rRNA transcription in previously untrained individuals.
This is supported by the observation that pre-rRNA increased rapidly initially in both protocols with minimal changes in response to subsequent sessions, regardless of exercise volume.
The CONST protocol in the present study corresponded to volumes used in the moderate volume condition in a previous study from our lab (three sets in two exercises activating knee extensor muscles) [@RN2358].
There, higher levels of total RNA were observed after four sessions in the moderate compared to a low volume protocol [@RN2358].
Interestingly, using a progressive volume protocol in well-trained participants, increases in total RNA have been reported throughout six weeks of training
[@RN2225].<!-- Haun total RNA increase througout a 6 week intervention -->
Although this observation was done in well-trained participants performing a high volume protocol without a control condition with constant volume, compared to constant volume protocols
[@RN1809;<!-- Brook 2016 -->
@RN2358],<!-- Hammarström 2020 -->
progressive volume may thus increase ribosomal abundance to a higher degree and provide a measure to avoid the plateau phase seen in the present study.

@@

## Muscle strength, body composition and muscle thickness assessments
Muscle strength was assessed as maximal voluntary isokinetic (90&deg; sec<sup>-1</sup>) and isometric (60&deg; angle, fully extended leg 0&deg;) knee extension torque. After a brief warm-up (5-min cycling, RPE 12-14), participants were seated and secured in the individually adjusted dynamometer. Participants were instructed to gradually increase their effort during three warm-up repetitions (50, 60 and, 70% of subjective maximal effort). After a 30-sec restitution period participants were instructed to perform three repetitions with maximal effort in the concentric phase. Sixty seconds after the isokinetic test the lever automatically moved to a 60&deg; angle and participants were instructed to push against the lever enough to see feedback from the visual feedback system. After an additional 15-sec restitution period, participants were instructed to push against the lever with maximal effort. Within the same assessment session, participants remained seated in the dynamometer for measurement performed on both legs. The first measurement was alternated between legs every other session. For statistical treatment of the data, all successful attempts were used. The last strength assessment at baseline was performed at least seven days prior to the first biopsy sampling. At least one of the baseline strength tests was performed on separate day with two sessions allowed to be perform on the same day with a short rest between assessments. Post training assessments were performed 48 hours and eight days after the last session.

@@

## Immunoblotting
Protein samples (20 &#956;g) were separated on 4-20% Tris-Glycin gels (Criterion TGX Precast Gels, Bio-Rad) at 250 V for 45 min using the recommended running buffer (25 mM Tris, 192 mM Glycin, 0.1% SDS). All samples from the same participant were run on the same gel and all samples were run in at least duplicates. When samples were compared between participants, signals were expressed per a calibrator sample from each participant that in turn was measured on a separate gel with all calibrator samples. Separated samples were transferred to &#956m PVDF membranes (Immun-Blot, Bio-Rad) using wet transfer (25 mM Tris, 192 Glycin, 10% vol/vol methanol) at a constant voltage of 300 mA for 3 h. Membranes were then stained to confirm transfer and enable total protein quantification using a reversible protein stain (Pierce Reversible Protein Stain, Thermo Fisher Scientiﬁc). Primary antibodies were acquired to detect UBF (F-9, sc-13125, Santa-Cruz Biotechnology, Dallas, Texas, USA) and ribosomal protein S6 (54D2, #2317, Cell Signaling Technology, Danvers, MA, USA). After blocking (Tris-buffered saline blocking buffer, 20 mM Tris, 150 mM NaCl, 5% fat-free milk, 0.1% Tween-20), membranes were incubated over-night with primary antibodies diluted in blocking buffer (UBF, 1:200; S6, 1:1000) followed by incubation with a secondary antibody conjugated to horseradish peroxidase (Anti-mouse IgG, #7076, Cell Signaling Technology, 1:10000). Membranes were washed 6 &#215; 5 min after incubation with primary antibodies and 8 &#215; 5 min after incubation with the secondary antibody. All incubation and washing steps were performed at 4&deg;C using an automatic membrane processor (BlotCycler, Precision Biosystems, Mansﬁeld, MA, USA). Chemiluminescent signals from membranes were detected after 5 min incubation in substrate (Super Signal West Femto Maximum Sensitivity Substrate, Thermo Fisher Scientiﬁc) using a documentation system. Total protein content was quantified from whole membrane images and defined as the mean gray value of the whole lane. Between-lane gray values were used as background subtracted from protein values. Total protein quantification was done using ImageJ Fiji [@RN2561]. Chemiluminescence signals were quantified using Image Studio Lite (LI-COR Biotechnology, Lincoln, NE, USA).
```{r}
library(tidyverse)

western_data <- readRDS("./data/derivedData/western-compile/western_data.RDS")

### CV of duplicate measurements

western_cv <- western_data %>%
  group_by(sample, target) %>%
  summarise(n = n(), 
            m = mean(expression, na.rm = TRUE), 
            s = sd(expression, na.rm = TRUE), 
            cv = s/m) %>%

  group_by(target) %>%
  summarise(cv = mean(cv, na.rm = TRUE)) %>%
  pull(cv)

cvs <- paste0(round(min(western_cv) * 100, 1), " and ", round(max(western_cv) * 100, 1), "%")
```
<span custom-style="CHANGE">The average coefficient of variation across replicates were `r cvs`, for RPS6 and UBF, respectively.</span>
